LDA exploration

### 2016-02-04 doc/manuscript.Rmd
Added to manuscript intro, result, and discussion seections

Title:

Author:

@@

Possible reason why IR is not shown... other studies have been cross-sectional..
Thus a potential confounder is the role that obesity plays in IR and IR on FFA
release... So, longitudinally, FFA would not predict IR since it is IR that is
suppressing NEFA.

Popular hypothesis: obese -> FFA -> IR.. may not be the case

total nefa evidence, well-studied. but there is diversity. no studies on it.

Some evidence to suggest that FFA delivery may be higher in obese, but that conc
may or may not be. Also that research on increasing FFA and IR comes from
experimental (probably fairly short term) research.

Association between higher FFA and muscle/hepatic IR, but challenging as adipose
tissue can also be IR. So is it the higher FFA -> IR or is it IR -> FFA?

methods -> gee, then lda (or similar) to confirm the role of 16:0 etc rather
than any individual FA

@@

THOUGHTS:

- sleep restriction can impact NEFA, but only higher during night and a bit
during early-morning. Should keep this in mind.
- Adjust for TAG in IR models?
- Key is that this is fasting, not postprandial!

@@

```{r setup, collapse=TRUE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = FALSE)
```

```{r loadData, message=FALSE}
source('.Rprofile')
suppressMessages(run_setup())
load_data()
dim(ds)
```

@@

sfig_consort <- sfig('consort', 'CONSORT diagram....')
sfig_lda <- sfig('lda', 'Linear discriminant factors with explained variance against tertiles of log(ISSI-2) in 463 PROMISE participants at the baseline visit (2004-2006).')
```

# Background #

Three paragraphs:

1. Total NEFA on diabetes, pathology, evidence
2. NEFA diverse, individual NEFA on pathology, bit more evidence from total, go
into experimental of species
3. Limitations, gaps, inconsistencies, objectives and hypotheses

# Subjects and Methods #

The Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort is a 
longitudinal observational study of participants with one or more risk factors 
for type 2 diabetes mellitus, including obesity, hypertension, family history of
diabetes, and/or a history of gestational diabetes or birth of a macrosomic 
infant.  Participants aged 30 years and older were recruited for the baseline 
visit from Toronto and London, Ontario, Canada into the PROMISE cohort (n=736) 
between 2004-2006.  Follow-up examinations occur every three years, with 
currently three examination visits completed to date (2004-2006, 2007-2009, and 
2010-2013).  The current study used data on participants who did not have 
diabetes at baseline, who returned for one or more follow-up examinations, and 
who had samples available for fatty acid measurements (n=477).  A diagram of the
sample size at each visit is shown in `r cite_sf('consort')`.  At each 
examination, participants undergo extensive metabolic characterization, 
anthropometric measurements, and structured questionnaires on lifestyle and 
sociodemographics.  Research ethics approval was obtained from Mount Sinai 
Hospital and the University of Western Ontario.  Data collection methods were 
standardized across the 2 centres and research nurses were centrally trained. 
<!--Over the 6-yrs, {{n return each visit}} returned for the 6-yr clinic visit 
({{pct loss to followup}}% loss to follow-up).  By the 6-yr visit, {{n converted
to IFG}} converted into IFG, {{n converted to IGT}} converted into IGT, and {{n 
converted to DM}} into diabetes.-->

## Blood measure assessment

At each examination, participants had an 8-12 hour fasting blood sample 
taken followed by a 75g oral glucose tolerance test (OGTT) with a 30 minute and
2 hour blood draw. All blood samples were processed and frozen at -70°C. {{Alanine
aminotransferase (ALT) was measured using standard laboratory procedures.}} 
Cholesterol, HDL, and triglycerides (TG) were measured using Roche Modular's
enzymatic colometric tests (Mississauga, ON).  Both specific insulin and glucose
were derived from the OGTT at fasting, 30 minute, and 2 hour time points. 
Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics, Basel, 
Switzerland) immunoassay analyzer and electrochemiluminescence immunoassay. 
This assay shows 0.05% cross-reactivity to intact human pro-insulin and the Des
31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.  Glucose
was determined using an enzymatic hexokinase (Roche Modular, Roche Diagnostics)
with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The inter-assay %CV is
<1.1% and intra-assay %CV is < 1.9%.  All assays were performed at the Banting 
and Best Diabetes Centre Core Lab at the University of Toronto. Impaired fasting
glucose (IFG), impaired glucose tolerance (IGT), and diabetes were categorized
using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the
baseline visit, which had been frozen at -70°C for 4-6 years and had not been
exposed to any freeze-thaw cycles.  Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount
of {{internal standard name}} (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch
[@Folch1957a].  Each serum lipid fraction was isolated using thin layer 
chromatography, which were  visualized under UV light after lightly spraying 
with 8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to
fatty acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1
h.  Fatty acid methyl esters were separated and quantified using a Varian-430
gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor
Four capillary column (VF-23ms; 30 m × 0.25 mm i.d. × 0.25 μm film thickness)
and a flame ionization detector, which were injected in splitless mode. Fatty
acid concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a;
@Chen2011a].  There were 22 fatty acids measured in the NEFA fraction.

@@

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Sociodemographic information, such as age, sex, and ethnicity, were determined
using questionnaires administered at each examination.

@@

The primary outcome variables for the PROMISE cohort and this analyses were 
1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this study were
the 22 NEFA as a mole percent (mol%) of the total fraction and as a 
concentration (nmol/mL).

Pearson correlation coefficients were computed for the PL and CE fatty acids
with other continuous variables. {{keep??}}

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables.  Given the longitudinal design
and auto-regressive (AR1) correlation matrix was chosen for the GEE models,
though other matrices (eg. exchangeable) had similar fit (data not shown)
{{confirm}}. GEE is well suited to cohort studies given it's capacity to handle
missed visits. The predictor variables and continuous covariates were scaled 
(mean centered and standardized). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables and covariates were set as *time-dependent*. 
No imputation was conducted on missing values.

Covariates selection was based on previous literature, directed acyclic graph
[@Greenland1999a] recommendations, and quasi-likelihood information criterion
(QIC). `r cite_sf('qic')` shows the covariates compared using QIC and 
`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model
(M#) we selected had time, sex, ethnicity, baseline age, waist circumference,
{{...}}, and family history of diabetes as covariates. <!--(exponentiated, minus
1, and multiplied by 100)--> After scaling, log-transforming, and exponentiating
the GEE estimates, the GEE results are interpreted as an expected percent
difference in the outcome variable for every SD increase in the predictor
variable given the covariates are held constant. An time by predictor term was
included in each of the GEE models to test for time interactions. Given that TG
is a risk factor for diabetes and that NEFA contribute to TG production, TG may
act as a mediator between NEFA and the outcomes. However, only a subset of the
participants at the 6-yr visit had clinically measured TG. To determine the role
of TG in the association between NEFA and the outcomes, TG was included in the
GEE model as a sensitivity analysis.

@@

```{r geeData, cache=TRUE}
outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
outcomes <- c(outcomes_is, outcomes_bcf)
ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT')
covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 
                    'MET', 'AlcoholPerWk')

gee_df <- get_gee_data(ds)
gee_results <- bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
    )

@@

```{r inlineGEE, dependson='geeData'}
# extract gee beta coefficient for indicating magnitude
pctChg.o <- ds %>%
  select(f.VN, HOMA, ISI, IGIIR, ISSI2) %>%
  gather(Measure, Value, -f.VN) %>%
  na.omit() %>%
  group_by(Measure, f.VN) %>%
  summarise(med = median(Value)) %>%
  spread(f.VN, med) %>%
  mutate(pctChg = ((`6-yrs` - `0-yr`) / `0-yr`) * 100) %>%
  select(pctChg) %>%
  abs() %>%
  round(1) %>%
  {paste0(min(.), '% to ', max(.), '%')}
```

```{r inlineLDA}
fit.class <- predict(lda_results, newdata = lda_df[, -1])$class %>% 
    factor(., ordered = TRUE)

# Amount of participants who were misclassified
misclass <- table(fit.class, lda_df[[1]]) %>% 
    as.data.frame() %>% 
    filter(fit.class != Var2) %>% 
    summarize(count = sum(Freq)) %>% 
    as.numeric()

# Amount of participants who were correctly classified
good_class <- table(fit.class, lda_df[[1]]) %>% 
    as.data.frame() %>% 
    filter(fit.class == Var2) %>% 
    summarize(count = sum(Freq)) %>% 
    as.numeric()

misclass_npct <- paste0(misclass, ' (', round(misclass / (good_class + misclass) * 100, 0), '%)')
```

```{r inlineMisc}
pctne <- calculate_percent_contribution(ds)

extractBeta <- function(data) {
    data %>%
      mutate(estCI = paste0('(beta: ', trim(format(round(estimate, 1), nsmall = 1)),
                            ' CI: ',
                            trim(format(round(conf.low, 1), nsmall = 1)),
                            ', ',
                            trim(format(round(conf.high, 1), nsmall = 1)),
                            ')'),
             #indep.estCI = paste(indep, estCI),
             #dep = paste(dep %>% gsub('log\\((.*)\\)', '\\1', .), estCI),
             dep = paste(dep %>% gsub('log\\((.*)\\)', '\\1', .))
             ) %>%
      filter(p.value <= 0.05) %>%
      arrange(dep, desc(estimate)) %>%
      select(dep, indep, estCI)
}
```

@@

## Distribution of NEFA

`r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four
individual NEFA made up the majority (`r sum(pct_nefa$pct)`%) of the total NEFA
fraction. The largest contributor was `r pctne$c[1]`, followed by `r
pctne$c[2]`, `r pctne$c[3]`, and `r pctne$c[4]`.

## Correlations of NEFA with basic characteristics

## GEE

As shown in {{supFigGeeUnadj}}, several associations were seen in the unadjusted
GEE models, particularly for the concentration data. Full model adjustment
attenuated most of these associations, with total NEFA, 16:0, and 18:2n-6 (all
as nmol/mL) had negative associations between IGI/IR and ISSI-2 ({{figGee}}).
The magnitude of association for each of these variables was fairly consistent
for each beta-cell function measure. For every one SD decrease in any of these
three NEFA variables, there was a predicted {{magnitude}}% lower IGI/IR and a
{{magnitude}}% lower ISSI-2 at each clinic visit. None of the NEFA variables as
a mol% had any association with the outcomes.

Adjusting for TAG attenuates all associations with IGI/IR and ISSI-2 (data not
shown). This may be due to either or both of two factors, mediation of TAG on
the pathway between NEFA and beta-cell function or the lower sample size at the
6 year visit from the incomplete measurement of TAG due to cuts in funding.

There were no significant interaction effects between time and the individual
NEFA on any of the outcome measures (data not shown).

@@

There was poor discriminatory ability in identifying patterns between tertiles
of ISSI-2, with only `r misclass_npct` of participants in the baseline
visit who were classified correctly. In general, participants in the 3rd tertile
(a high ISSI-2) had less of all NEFA, though the mean differences between
tertiles were minor (`r cite_f('lda')`). The separation of tertiles of
log(ISSI-2) and linear discriminants is shown in `r cite_sf('lda')`.

Results between visits were not substantially different.

@@

We found that higher palmitic acid (16:0) and linoleic acid (18:2n-6) associated
with lower beta-cell function by {{number?}}. Expand on experimental studies,
pathology/biology, mechanisms, etc.

There is substantial experimental research on the role of increasing NEFA and
increasing IR {{cite}}. Two major limitation of these studies are that they tend
to be fairly short term ({{length of time}}) and they tend to increase NEFA to
supraphysiological concentrations {{confirm}}. Epidemiological research has also
shown a link between higher NEFA and higher IR {{cite}}. However, most of these
studies are cross-sectional in nature and can not address causal direction.
Adipose tissue can become insulin resistant and thus hormone-sensitive lipase
may not respond to the inhibitory action of insulin {{cite}}. As such, IR may
temporally occur sooner than elevations in blood NEFA. In this longitudinal
analysis, we saw no association between NEFA composition and neither hepatic IR
(HOMA-IR) nor whole-body insulin sensitivity (ISI). Possible explanations for
this could be that over several years NEFA does not influence hepatic or muscle
IR or that it is IR that elevates NEFA and as such NEFA would not predict IR
over time.

In the sensitivity analysis adjusting for TAG, all associations with beta-cell
function were attenuated. While there is strong experimental evidence on the
role of chronically elevated NEFA, particularly palmitic acid, on impaired
beta-cell function, within a free-living population NEFA may be mediating it's
association through TAG. In normal metabolism, NEFA enters the liver and assists
in the production of TAG that is to be processed into very-low density
lipoproteins (VLDL) {{cite}}. Elevated NEFA may be mediating it's association
through hypertriglyceridemia, a known risk factor for diabetes {{cite}}. A major
caveat with the TAG-adjusted GEE models is the reduced sample size in the 6 year
visit, as only a subset (n={{sample}}) of participants had TAG measured.

@@

# References #

@@

```{r figGEE, dependson='geeData', fig.cap=fig_gee}
plot_gee_results(gee_results)
```

@@

```{r supFigGeeUnadj, dependson='geeData'}
plot_gee_results(gee_results_unadj)
```

>>>>

Title: Total non-esterified fatty acids, but not composition, predict
progression in beta-cell dysfunction

Author: Luke W. Johnston, Richard P. Bazinet, Anthony J. Hanley

@@

@@

TODO:
- Adjust for baseline prediabetes? SenAnalysis?
- Do PLS-LDA instead of LDA, since it is apparently better for this type of
analysis/data

@@

```{r knit_setup, collapse=TRUE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = FALSE)
```

```{r setup, message=FALSE}
source('.Rprofile')
suppressMessages(run_setup())
load_data()
ds <- ds %>% 
    mutate(lALT = log(ALT),
           lTAG = log(TAG))
dim(ds)
```

@@

sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006)')
sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. Only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values presented were adjusted for the false discovery rate.')
sfig_lda <- sfig('lda', 'Linear discriminant factors with explained variance against tertiles of log(ISSI-2) in 463 PROMISE participants at the baseline visit (2004-2006).')
```

# Background

Total non-esterified fatty acids (NEFA) have been well documented to influence
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that
exposure of cell or animal to high concentrations of NEFA can induce insulin
resistance in insulin-sensitive tissue such as the muscle or liver, impair
beta-cell production of insulin, and cause hyperglycemia {{confirm...cite}}.
Epidemiological and clinical studies have observed similar findings, showing
that elevated total plasma NEFA associates with risk for diabetes {{cite}}.

Most previous experimental work uses palmitic acid (16:0), oleic acid (18:1n-9),
or a standard oil such as soybean {{confirm, cite}} and generalize as total
NEFA. However, fatty acids comprise multiple classes of molecules with diverse 
physiological functions, with few studies comparing different fatty acids. For
instance, one study on beta-cell {{plates? lines?}} suggests that
eicosapentaenoic acid (20:5n-3) can protect against the lipotoxic effect of
palmitic acid {{xiao? cite}}.

Prospective with diabetes [@Ilyasova2010a]

Lifestyle interventions aimed at managing or preventing diabetes show a
concomitant reduction in NEFA and insulin resistance {{cite}}.

Lack of longitudinal data is particularly important given that NEFA release from
adipocytes is inhibited by insulin, so if there is IR, more NEFA are released,
making it difficult to disentangle which caused which.

Despite the fairly extensive literature on NEFA and diabetes, there are
substantial research gaps. Importantly, most of the experimental work using
individual NEFA species has been in animal and cell lines or if in humans has
been acute, infusion studies. For epidemiological studies, few have examined the
role of total, but not individual, NEFA on the pathogenesis of diabetes and
fewer still have been longitudinal, which is key to understanding causal
associations and given the long timescale in diabetes pathophysiology.
Presently, no longitudinal study has been conducted that has examined the role
of the composition of the NEFA fraction on the pathogenesis of diabetes. Our
research objective was to examine the longitudinal role of the NEFA composition
on insulin sensitivity and beta-cell function. Some specific hypotheses included
that higher palmitic acid while lower polyunsaturated fatty acids such as
eicosapentaenoic acid would associate with lower insulin sensitivity and
beta-cell function.

# Subjects and Methods

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE included having one or more risk factors for type 2
diabetes mellitus, such as obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=726) conducted the baseline visit
between 2004-2006. Follow-up examinations occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
current study used data on participants who did not have diabetes at baseline,
who returned for one or more follow-up examinations, and who had samples
available for fatty acid measurements (n=477).  A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`.  At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics.  Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario. 
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

## Blood measure assessments

At each examination, an 8-12 hour fasting blood sample was draw from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triglycerides (TG) were measured using Roche
Modular's enzymatic colometric tests (Mississauga, ON).  Both specific insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at the University of
Toronto. Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and
diabetes were categorized using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles.  Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) {{confirm}} was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a].  Each serum lipid fraction was isolated using thin layer 
chromatography, which were visualized under UV light after lightly spraying with
8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to fatty
acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1 h. 
Fatty acid methyl esters were separated and quantified using a Varian-430 gas
chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor Four
capillary column and a flame ionization detector, which were injected in
splitless mode. Fatty acid concentrations (nmol/ml) were calculated by
proportional comparison of gas chromatography peak areas to that of the internal
standards [@Nishi2014a; @Chen2011a].  There were 22 fatty acids measured in the
NEFA fraction.

@@

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Sociodemographic information, such as age, sex, and ethnicity, were determined
using questionnaires administered at each examination. In the lifestyle
questionnaire, physical activity was determined using a version of the
Modifiable Activity Questionnaire (MAQ) [@Kriska1990].  The MAQ collects 
information on leisure and occupational activity, including intensity, 
frequency, and duration, over the past year.  Each reported activity from the 
MAQ was weighted by its metabolic intensity allowing for the estimation of 
MET-hours per week.

@@

The primary outcome variables for the PROMISE cohort and this analyses were 
1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this study were
the 22 NEFA as a mole percent (mol%) of the total fraction and as a 
concentration (nmol/mL). Pearson correlation coefficients were computed for the
individual NEFA and with other continuous variables.

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables.  Given the longitudinal design 
and auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to cohort studies given it's capacity to handle missed visits.
The predictor variables and continuous covariates were scaled (mean centered and
standardized). The NEFA variables were classified as *time-independent* (held
constant) as they were measured only at the baseline visit, while the outcome
variables and covariates were set as *time-dependent*. No imputation was
conducted on missing values.

Covariates selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and `r 
cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last few models were comparable in performance, so we chose the
simplest model (M8), adjusting for time, sex, ethnicity, baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While models with a
time interaction did not perform better than others for the QIC (M6.1 and M8.1
in `r cite_st('qic')`, we ran models with a time by predictor term for each of
the NEFA to test if there was a time interaction. Given that TG is a risk factor
for diabetes and that NEFA contribute to TG production, TG may act as a mediator
between NEFA and the outcomes. However, only a subset of the participants at the
6-yr visit had clinically measured TG. To determine the role of TG in the
association between NEFA and the outcomes, TG was included in the GEE model as a
sensitivity analysis.

@@

```{r varSet}
outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
outcomes <- c(outcomes_is, outcomes_bcf)
ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'MET', 'AlcoholPerWk')
covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')

gee_df <- get_gee_data(ds)
```

```{r geeData, cache=TRUE}
gee_results <- bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
    ) 
gee_results %>% str
# Use substr after this above to make the Xterms go in order.

@@

```{r inlineResults}
pctne <- calculate_percent_nefa_contribution(ds) 
misclass_npct <- calculate_misclass_npct(lda_results, lda_df)
pctChg.o <- calculate_outcomes_pct_change(ds)

@@

The primary outcome variables declined between `r pctChg.o` over the 6 years.

`r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four 
individual NEFA made up the vast majority (`r sum(pct_nefa$pct)`%) of the total
NEFA fraction. The largest contributor was `r pctne$c[1]`, followed by
`r pctne$c[2]`, `r pctne$c[3]`, and `r pctne$c[4]`.

All individual NEFA as well as the total fraction had weak to no correlations
(r<0.3) with the basic characteristics (see `r cite_sf('corr')`).

## GEE modeling

Several associations were seen in the unadjusted GEE models, particularly for 
the concentration data. Full model adjustment attenuated most of these 
associations, with total NEFA, 16:0, and 18:2n-6 (all as nmol/mL) had negative 
associations between IGI/IR and ISSI-2 (see `r cite_f('gee')`). The magnitude of
association for each of these variables was fairly consistent for each beta-cell
function measure. For every one SD decrease in any of these three NEFA 
variables, there was a predicted {{magnitude}}% lower IGI/IR and a 
{{magnitude}}% lower ISSI-2 at each clinic visit. None of the NEFA variables as 
a mol% had any association with the outcomes. Adjusting for TG attenuates all 
associations with IGI/IR and ISSI-2 (data not shown), though sample size at the 
6 year visit is restricted to n=128 in those who had TG measured. There were no
significant interaction effects between time and the individual NEFA on any of
the outcome measures (data not shown).

@@

There was poor discriminatory ability in identifying patterns between tertiles 
of ISSI-2, with only `r misclass_npct` of participants in the baseline visit who
were classified correctly. In general, participants in the 3rd tertile (a high
ISSI-2) had less of all NEFA, though the mean differences between tertiles were
minor (`r cite_f('lda')`). The separation of tertiles of log(ISSI-2) by the two
linear discriminants is shown in `r cite_sf('lda')`. Analyses on the later
visits to confirm the baseline results were not substantially different from the
baseline visit.

@@

## Beta cell

We found that higher palmitic acid (16:0), oleic acid (18:1n-9), and linoleic
acid (18:2n-6) associated with lower beta-cell function by {{number?}}. Expand
on experimental studies, pathology/biology, mechanisms, etc.
[@Salgin2012a;@Giacca2011a;@Xiao2009a]

Epidemiology [@Johns2014a]

Experimental [@Carpentier2003a]
Review [@Carpentier2000a;@Cnop2008]

Most experimental studies on beta-cell function use only palmitic acid....

First study to our knowledge to examine a broad spectrum of individual NEFA in a cohort setting on beta-cell function..

We found no association of any individual NEFA as mol% on beta-cell function,
but a very strong signal with total NEFA, palmitic acid, oleic acid, and
linoleic acid as concentrations.

## IR

There is substantial experimental evidence highlighting the role of increased 
NEFA and a subsequent increase in insulin resistance via impairment of insulin 
signaling cascades (reviewed in multiple articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a;@Surampudi2009]). However, in this
longitudinal analysis, we saw no association between any individual or total
NEFA and neither hepatic (1/HOMA-IR) nor whole-body insulin sensitivity (ISI). 
There are some possible explanations for our findings. Previous studies done *in
vivo* were mostly short term infusion 
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and the observational research was generally
cross-sectional [@Miller2012a;@Johns2014a;@Bush2012a]. The limitation of these
studies were the short time periods examined and the experimentally-induced
elevations in NEFA concentrations, making it difficult to assess the causal role
in non-experimental environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a], suggesting that either NEFA only 
have a minor role in IR or that NEFA is the effect, but not cause, of the 
underlying IR. Adipose tissue, which release NEFA during fasting, can become 
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA update [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA is a consequence of higher IR.

Another explanation for our null findings with insulin sensitivity may be
differences in physiology between fasting and postprandial NEFA kinetics. For
instance, some experimental studies using the clamp found that fasting NEFA was
a weak predictor of insulin sensitivity compared to postprandial concentrations
of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the
adipose tissue following postprandial TAG lipolysis via lipoprotein lipase
may result in NEFA spillover into the blood and a subsequent increase in circulating
NEFA [@Almandoz2013a], which may be more metabolically active given postprandial
activity.

...Last possibility is that given these subjects were at-risk for diabetes already,
the causal role of NEFA on IR may have long passed and it no longer influences
IR....

In the sensitivity analysis adjusting for TAG, all associations with beta-cell 
function were attenuated. While there is strong experimental evidence on the 
role of chronically elevated NEFA, particularly palmitic acid, on impaired 
beta-cell function, within a free-living population NEFA may be mediating it's 
association through TAG. In normal metabolism, NEFA enters the liver and assists
in the production of TAG that is to be processed into very-low density 
lipoproteins (VLDL) [@Nielsen2012a]. Elevated NEFA may be mediating it's
association through hypertriglyceridemia, a known risk factor for diabetes.
However, a major limitation to the TAG-adjusted GEE models is the reduced sample
size in the 6 year visit, as only a subset (n=128) of participants had TAG
measured.

## Composition

This is the first study to our knowledge to examine the composition of the NEFA fraction and

Composition of NEFA and metabolic syndrome components [@Dai2015a]

Composition of NEFA and metabolic syndrome binary [@Novgorodtseva2011a]

Composition of NEFA and diabetes status (x-sec) [@Liu2010a], found 16:0, 18:0,
18:1, 18:2, 18:3 could be useful biomarkers for identifying healthy vs diabetes

@@

@@

```{r figGEE, dependson='geeData', fig.cap=fig_gee}
gee_results[sort(order(
    gee_results$Xterms)[
        gee_df %>%
    select(matches('ne\\d\\d')) %>%
    gather(FA, Conc) %>%
    transmute(FA = renaming_fa(FA)) %>%
    unique() %>% 
        .[[1]]
    ])]
order(gee_results$Xterms)
plot_gee_results(gee_results)
gee_results %>% filter(Xterms == 'Total')
gee_results %>% str
```

@@

![`r sfig_consort`](img/flowDiagramSample.png)

```{r supTabQIC}
prep_qic <- gee_df %>%
    select(
        SID, VN, lISSI2, lISI, ne_Total, Sex, Ethnicity, MET, ALT, BaseAge,
        Waist, BMI, AlcoholPerWk, TobaccoUse, FamHistDiab
        )

qic_is <- prep_qic %>% 
    select(-lISSI2) %>% 
    na.omit()

qic_bcf <- prep_qic %>% 
    select(-lISI) %>% 
    na.omit()

# QIC for IS
M0 <- geepack::geeglm(
    lISI ~ ne_Total + VN, data = qic_is,
    id = SID, family = gaussian,
    corstr = 'ar1'
    )

M1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
M2 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT)
M3 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
M4 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk) # Best
M5 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # Best
M6 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # Best
M6.1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse + ne_Total:VN) # Best
